---
title: "Module 1: Vertical Multimodal Integration using R"
author: "Saptarshi Roy, Texas A&M University"
date: "ISMB 2026, Washington DC" 
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    theme: readable
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    output_dir: '.'
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

> **ISMB 2026 Tutorial VT4, Multimodal Integration and Multimodal Causal Inference using R/Bioconductor.** This is Module 1 of three hands-on modules delivered in a single day. Module 1 covers vertical (multi-omics) integration, Module 2 covers horizontal (meta-analysis) integration, and Module 3 covers multiview mediation with Zentangler. Before starting, run the setup script once (`Rscript setup/install.R`) to install and load every package used across all three modules.

# Setup

Install the following R packages and set a random seed to get started.

```{r pkgs, message = FALSE, warning=FALSE}
## Load the following R packages
### Global packages
library(tidyverse) 
library(ggplot2) 
library(cowplot) 
library(bayesplot) 
library(dplyr) 
library(ROCR)
library(purrr)

### For IntegratedLearner implementation
library(caret) 
library(SuperLearner) 
options(java.parameters = "-Xmx8000m")
gc()
library(bartMachine)
library(bartMachineJARs)
library(IntegratedLearner) 

### For JAFAR implementation
library(jafar) 

### For multiview implementation
library(glmnet) 
library(multiview) 

### For BayesCOOP implementation
library(BayesCOOP) 

## Set a seed for reproducibility
set.seed(123)
```

In this lab, we showcase examples of various integration paradigms (early, late, and intermediate fusion) using the R packages [`multiview`](https://cran.r-project.org/web/packages/multiview/index.html),
[`IntegratedLearner`](https://github.com/himelmallick/IntegratedLearner),
[`BayesCOOP`](https://github.com/himelmallick/BayesCOOP) and
[`jafar`](https://github.com/niccoloanceschi/jafar).

# Pregnancy Dataset (Continuous Response)

We load the pregnancy dataset from the [IntegratedLearner GitHub
repository](https://github.com/himelmallick/IntegratedLearner). This dataset has an independent **validation** set to facilitate model
development and evaluation.

In this dataset, we focus on predicting **time-to-labor onset** using multi-omics measurements from the immunome, metabolome, and proteome for a cohort of women who experienced spontaneous labor. The dataset, described in [Mallick et al. (2024)](https://pubmed.ncbi.nlm.nih.gov/38146838/), consists of repeated blood measurements collected during the final 100 days of pregnancy.

To align with prior work (e.g., [Ding et al., 2022](https://www.pnas.org/doi/10.1073/pnas.2202113119)) and to simplify the modeling framework, we construct a cross-sectional sub-dataset by retaining only the first available measurement for each participant. We treat \underline{time-to-labor onset as a continuous outcome}.

The resulting dataset represents an extreme *large-p-small-n* setting. The three molecular layers provide high-dimensional predictors with  
- \( p_1 = 1141 \) single-cell immune features,  
- \( p_2 = 3529 \) metabolites, and  
- \( p_3 = 1317 \) proteins,  

for a total of \( n_{\text{tot}} = 53 \) subjects. Similarly, we have $8$ subjects in the validation set with data on metabolite layer missing. 

This high-dimensional structure motivates the use of integrative and regularized modeling approaches capable of borrowing strength across modalities while controlling overfitting.

Each of these datasets comes with a structured list with three components that
together represent a complete multi-omics (or multi-view) dataset:

-   **`feature_table`** – A matrix of quantitative features such as
    microbial taxa abundances, metabolite intensities, or other
    high-dimensional predictors. In this representation, **columns
    correspond to samples** and **rows correspond to features**. This
    layout is compatible with the downstream modeling functions used in
    this workflow.

-   **`feature_metadata`** – A data frame of annotations that describe
    each feature in `feature_table`. Typical metadata may include the
    data modality (e.g., taxonomic group, biochemical category), pathway
    membership, measurement platform, or other biological or technical
    descriptors. This metadata enables view-specific analyses and
    interpretation of model outputs.

-   **`sample_metadata`** – A data frame of sample-level information
    that accompanies `feature_table`. This typically includes the
    outcome or phenotype of interest (e.g., clinical status), subject
    identifiers, covariates such as age or batch variables, and any
    grouping or stratification variables relevant for training and
    validation. This component is essential for supervised modeling and
    performance assessment.

```{r load_data}
## Load train data
data_train <- get(load(url(
  "https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS.RData"
)))

## Load test data
data_test <- get(load(url(
  "https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/StelzerDOS_valid.RData"
)))

## How many layers and how many features per layer?
table(data_train$feature_metadata$featureType)

## Explore data dimension of feature_table: no. of features x no. of samples
print(dim(data_train$feature_table))

## How do the components look like?
print(data_train$feature_metadata[c(1:5, 1441:1445, 2758:2762), ])
head(data_train$feature_table[1:5, 1:5])
head(data_train$sample_metadata[1:5, ])
```

## Preprocessing the Dataset

For simplicity, we remove **Metabolomics** features because they are not
present in the validation set. We then realign the feature tables with
the filtered metadata so that only valid rows and columns remain.
Finally, we restrict the analysis to baseline samples only, selecting
training samples labeled “A” and test samples labeled “G1.” This ensures
that both the feature matrices and the sample metadata remain perfectly
synchronized across datasets.

**NOTE**: Although we take this step for the purpose of this tutorial,
it is not strictly necessary. The [IntegratedLearner
paper](https://pubmed.ncbi.nlm.nih.gov/38146838/) includes a case study
demonstrating how to use all available data effectively, even when one
modality is completely missing in the validation set. This example was
also discussed during the lecture.

```{r preprocess_data}
## Removing Metabolomics
data_train$feature_metadata <- data_train$feature_metadata %>%
  filter(featureType != "Metabolomics")

data_train$feature_table <- data_train$feature_table[
  rownames(data_train$feature_metadata), , drop = FALSE
]

## Selecting baseline observations from train data
pos_train <- grep("A", colnames(data_train$feature_table), ignore.case = TRUE)
data_train$feature_table  <- data_train$feature_table[, pos_train, drop = FALSE]
data_train$sample_metadata <- data_train$sample_metadata[pos_train, , drop = FALSE]

## Selecting baseline observations from test data
pos_test <- grep("G1", colnames(data_test$feature_table))
data_test$feature_table   <- data_test$feature_table[, pos_test, drop = FALSE]
data_test$sample_metadata <- data_test$sample_metadata[pos_test, , drop = FALSE]
```

To reduce noise and remove uninformative features, we filtered the data
using predefined **abundance** and **prevalence** thresholds. Abundance
filtering removes consistently negligible features, while prevalence
filtering excludes features present in only a small fraction of samples.

In our analysis, we set the abundance threshold to $0$, meaning no
features were removed based on magnitude, and applied a prevalence
threshold of $0.1$, retaining features observed in at least $10\%$ of
samples. This reduces instability from extremely rare features while
improving interpretability and computational efficiency.

We will use three internal functions, `gen_datalist()`,
`reconstruct_data()` and `filter_features()`, from the `BayesCOOP`
package to carry out this preprocessing step. These functions are not
exported but are designed to streamline data preparation within the
modeling workflow. A brief description of each function is given below:

-   `gen_datalist()`: This function takes as input a structured list
    containing `feature_table`, `feature_metadata`, and
    `sample_metadata`, and transforms it into the format required for
    multi-view modeling. In particular, it partitions the feature table
    according to predefined data views and returns an object where
    `xList` is a list of feature matrices, with each element
    corresponding to one data modality (or view). Each matrix in `xList`
    contains the features for that specific view, aligned consistently
    across samples. The function also retains the associated response
    and metadata, ensuring that all components are properly synchronized
    for downstream model fitting.

-   `reconstruct_data()`: This function performs the inverse
    transformation of `gen_datalist()`. It reconstructs the processed
    data back into the original structured format (with `feature_table`,
    `feature_metadata`, and `sample_metadata`) after intermediate
    transformations, facilitating interpretation and further analysis.

-   `filter_features()`: This function performs feature-level filtering
    based on abundance and prevalence criteria. Given a feature matrix
    `x` (with samples in rows and features in columns), it first retains
    only those features whose values exceed a specified abundance
    threshold in more than a given proportion of samples (the prevalence
    threshold). This removes rare or sparsely observed features that may
    introduce noise or instability in downstream modeling. It then
    applies `caret::nearZeroVar()` to eliminate features with near-zero
    variance, further reducing uninformative predictors. The function
    returns the filtered feature matrix.

```{r}
## Filtering features according to abundance and prevalence threshold
xList_train <- BayesCOOP:::gen_datalist(data_train)$xList
y_train <- BayesCOOP:::gen_datalist(data_train)$y
xList_test <- BayesCOOP:::gen_datalist(data_test)$xList
y_test <- BayesCOOP:::gen_datalist(data_test)$y
abd_thresh <- 0; prev_thresh <- 0.1
xList_train <- lapply(xList_train, function(foo) BayesCOOP:::filter_features(foo, abd_thresh, prev_thresh))
xList_test <- lapply(xList_test, function(foo) BayesCOOP:::filter_features(foo, abd_thresh, prev_thresh))

## Considering the common features between train and test set
keep_features <- vector("list", length = length(xList_train))
for(i in 1:length(keep_features)) {
  keep_features[[i]] <- intersect(names(xList_train[[i]]), names(xList_test[[i]]))
  xList_train[[i]] <- as.matrix(xList_train[[i]][, keep_features[[i]], drop = FALSE])
  xList_test[[i]] <- as.matrix(xList_test[[i]][, keep_features[[i]], drop = FALSE])
}

## Reconstructing the train and testing data object
data_train_processed <- BayesCOOP:::reconstruct_data(xList_train, y_train)
data_test_processed <- BayesCOOP:::reconstruct_data(xList_test, y_test)

## Change column name to make it uniform across all packages henceforth
colnames(data_train_processed$sample_metadata)[1] <- "subjectID"
colnames(data_train_processed$feature_metadata) <- c("featureID", "featureType")
colnames(data_test_processed$sample_metadata)[1] <- "subjectID"
colnames(data_test_processed$feature_metadata) <- c("featureID", "featureType")

## Check new dimensions after filtering of features
print(dim(data_train_processed$feature_table))
print(dim(data_test_processed$feature_table))
```

## Integrated Learner using `IntegratedLearner` R Package

`IntegratedLearner` late fusion algorithm proceeds by 1) fitting a
machine learning algorithm per-layer to predict outcome (`base_learner`)
and 2) combining the layer-wise cross-validated predictions using a meta
model (`meta_learner`) to generate final predictions based on all
available data points. As a default choice, we recommend `SL.nnls.auc`
as the meta model algorithm. It fits a non-negative least squares (in
case of continuous outcome) and rank loss minimization (in case of
binary outcome) on layer-wise cross-validated predictions to generate
the final predictions and quantify per-layer contribution in the final
predictions. For further details, see [the
paper](https://pubmed.ncbi.nlm.nih.gov/38146838/).

The default base model recommendation for `IntegratedLearner` is
Bayesian additive regression trees or BART (`base_learner='SL.BART'`).
Using BART as base-model yields uncertainty estimates (i.e. credible
intervals) of the prediction and model parameters in addition to
reporting a small set of interpretable features for follow-up
experiments (i.e. feature importance scores).

Below we briefly describe the main arguments controlling model fitting
and integration in `IntegratedLearner()`:

-   **`base_learner`**\
    The algorithm used to fit a model for each data view. Each view is
    modeled separately before integration. The default `"SL.BART"`
    provides uncertainty estimates, but here we use `"SL.randomForest"`
    for speed.

-   **`base_screener`**\
    Controls whether features are pre-filtered before fitting each base
    learner. `"All"` (default) uses all features with no screening.

-   **`meta_learner`**\
    The algorithm that combines base learner predictions across views in
    the stacked model. The default `"SL.nnls.auc"` fits a constrained
    least squares or rank-loss model to yield final integrated
    predictions.

-   **`run_concat`**\
    If `TRUE`, fits an early fusion model by concatenating all views
    into one learner.

-   **`run_stacked`**\
    If `TRUE`, runs the stacked late-fusion model with the specified
    `meta_learner`.

-   **`refit.stack`**\
    When `TRUE`, refits the stacked model on the full training data
    after cross-validation; `FALSE` returns predictions based on the
    original CV fits.

-   **`verbose` / `print_learner`**\
    Control console output during fitting: `verbose` shows progress
    messages, `print_learner` prints the learner summary.

```{r IL, message=FALSE, warning=FALSE}
## Fit IntegratedLearner
fit_IntegratedLearner <- IntegratedLearner(
      feature_table = data_train_processed$feature_table,
      sample_metadata = data_train_processed$sample_metadata,
      feature_metadata = data_train_processed$feature_metadata,
      feature_table_valid = data_test_processed$feature_table,
      sample_metadata_valid = data_test_processed$sample_metadata,
      folds = 5,
      seed = 123,
      base_learner = "SL.randomForest",
      base_screener = "All",
      meta_learner = "SL.nnls.auc",
      run_concat = TRUE,
      run_stacked = TRUE,
      verbose = FALSE,
      print_learner = TRUE,
      refit.stack = TRUE,
      family = gaussian()
    )
```

### Prediction performance

Below we evaluate predictive accuracy on the held-out test set using
both early fusion and late (stacked) fusion results from
`IntegratedLearner`. We compute the predicted values of the outcome and
then quantify performance using mean squared prediction error (MSPE),
which measures the average squared difference between predicted and true
values. Lower MSPE indicates better predictive performance.

```{r, IL_pp}
## Prediction values of Y for both early and late fusion schemes
y_pred_early <- fit_IntegratedLearner$yhat.test$concatenated
y_pred_late <- fit_IntegratedLearner$yhat.test$stacked

## Calculation of mean squared prediction error
mspe_IL_early <- mean((y_pred_early - y_test)^2)
print(mspe_IL_early)
mspe_IL_late <- mean((y_pred_late - y_test)^2)
print(mspe_IL_late)
```

## Cooperative Learning using `multiview` R Package

The `multiview` package implements a supervised learning framework known
as *cooperative learning* for multi-view data integration. In this
setting, the goal is to learn a predictive model from multiple feature
sets (or “views”) that may represent different data modalities (e.g.,
taxa abundances, metabolite levels, clinical measures). Rather than
fitting each view independently or naïvely concatenating all views,
cooperative learning balances two objectives: (1) minimizing the
prediction loss (such as squared error for continuous outcomes or
deviance for generalized models), and (2) encouraging the predicted
responses from different views to *agree* with each other.

To achieve this, the method introduces an *agreement penalty* whose
strength is controlled by a tuning parameter $\rho$. When $\rho$ is
small, the views are fit more independently (similar to early fusion).
As $\rho$ increases, the views are forced to produce more similar
predictions, leading toward a late fusion (or consensus) solution. By
varying $\rho$, the method traces a continuum of models that interpolate
between independent and fully cooperative fits, allowing the data to
determine the degree of agreement that best balances individual view
performance and cross-view coherence.

In addition, cooperative learning integrates a *sparsity-inducing
penalty* (ridge, lasso and elastic net) with the agreement penalty. This
combined penalty not only promotes agreement across views but also
yields interpretable models by selecting a subset of relevant features
from each view. The resulting framework is flexible and well-suited for
high-dimensional multi-view settings where views may contain redundant
or noisy predictors.

If we assume that we have `M` data matrices
$X_1 \in \mathbb R^{n\times p_1},X_2\in \mathbb R^{n\times p_2},\ldots, X_M\in \mathbb R^{n\times p_M}$,
`multiview` essentially solves the problem
$$\frac{1}{2} ||y- \sum_{m=1}^{M} X_m\rho_m||^2+  \frac{\rho}{2}\sum_{m<m'} ||(X_m\rho_m- X_{m'}\rho_{m'})||^2 + \sum_{m=1}^{M} \lambda_m \Bigl[(1-\alpha)\|\rho_m||_1 + \alpha||\rho_m||_2^2/2\Bigr].$$

We will be using the following two functions to run `multiview` workflow
for different fusion schemes.

-   `cvfit_alpha()`: A helper function that runs **cross-validation**
    for the `multiview` model to select the optimal regularization
    parameter (`lambda`) for a given agreement penalty (`rho`). It takes
    a list of view-specific feature matrices (`xlist`), the outcome
    vector `y`, the agreement parameter `rho`, and optional arguments
    for the elastic net mixing parameter (`alpha`) and number of folds.
    The function returns the value of `lambda` that minimizes the
    cross-validated loss, which is used in downstream model fitting.

-   `fit_multiview()`: A wrapper function that encapsulates the workflow
    of training, evaluating, and extracting results from a `multiview`
    cooperative learning model for a specific fusion scheme. Given train
    and test data (feature lists and outcome vectors) and a fusion
    parameter `rho`, it first labels the fusion scheme (Early /
    Intermediate / Late), selects the best tuning parameter using
    `cvfit_alpha()`, and fits the `multiview` model. It then generates
    predictions on the test set and computes **mean squared prediction
    error (MSPE)**. The function also extracts the model coefficients,
    counts the number of selected (non-zero) features (excluding the
    intercept), and returns a tidy data frame of non-zero coefficients
    with associated feature names, view labels, and fusion scheme
    identifiers. The output is a list containing the fitted model
    object, predictions, MSPE, selected feature count, and the non-zero
    coefficient table for downstream analyses or visualization.

```{r coop}
## Function to choose lambda using cross validation
cvfit_alpha <- function(xlist, y, rho, alpha = 1, folds = 5) {
  cvfit <- multiview::cv.multiview(x_list = xlist, y = y, rho = rho, alpha = alpha, nfolds = folds)
  list(lambda_min = cvfit$lambda.min)
}

## Function that outlines the workflow to 
## run multiview for different fusion schemes
fit_multiview <- function(xList_train, y_train, xList_test, y_test, rho, alpha = 1){
  # Extracting the fusion scheme
  if(rho == 0)
    scheme_label = "Early"
  else if(rho == 0.5)
    scheme_label = "Intermediate"
  else
    scheme_label = "Late"
  
  # Cross-validate to choose best lambda
  cvres      <- cvfit_alpha(xList_train, y_train, rho = rho, alpha = alpha)
  best_lambda <- cvres$lambda_min

  # Fit multiview at chosen lambda
  fit_mv <- multiview::multiview(
    xList_train, y_train,
    lambda = best_lambda,
    rho    = rho,
    alpha  = 1
  )
  
  # Predictions on the test set
  y_pred_mv <- predict(
    fit_mv,
    newx  = xList_test,
    s     = best_lambda,
    rho   = rho,
    alpha = 1
  )

  # Compute MSPE
  mspe_mv <- mean((y_pred_mv - y_test)^2)

  # Extract coefficients
  coefs_mv <- coef(
    fit_mv,
    s     = best_lambda,
    rho   = rho,
    alpha = 1
  )
  
  # Count nonzero (exclude intercept)
  n_nonzero_mv <- sum(coefs_mv != 0) - 1

  # Tidy into a data frame
  coef_df <- as.data.frame(as.matrix(coefs_mv))
  coef_df$feature <- rownames(coef_df)

  nonzero_coefs <- coef_df %>%
  tidyr::pivot_longer(
    cols     = -feature,
    names_to = "coef_column",
    values_to = "coefficient"
  ) %>%
  dplyr::filter(coefficient != 0, feature != "(Intercept)") %>%
  dplyr::mutate(
    # Map feature to its view (e.g., proteomics, CyTOF)
    view = data_train_processed$feature_metadata$featureType[
      match(feature, data_train_processed$feature_metadata$featureID)
    ],
    method = scheme_label,
    rho    = rho
  ) %>%
  dplyr::select(feature, view, coefficient, method, rho)

  return(list(fit_mv = fit_mv, y_pred_mv = y_pred_mv, mspe_mv = mspe_mv, n_nonzero_mv = n_nonzero_mv, nonzero_coefs = nonzero_coefs))
}

## Different rho values for choice of fusion schemes
rhos_to_try <- c("Early" = 0, "Intermediate" = 0.5, "Late" = 1)
```

```{r}
set.seed(123)
## Matrix datatype conversion
xList_train <- lapply(xList_train, as.matrix)
xList_test <- lapply(xList_test, as.matrix)

## Run multiview for Early fusion
fit_mv_early <- fit_multiview(xList_train, y_train, xList_test, y_test, rho = 0)

## Run multiview for Late Fusion
fit_mv_late <- fit_multiview(xList_train, y_train, xList_test, y_test, rho = 1)

## Run multiview for Intermediate Fusion
fit_mv_intermediate <- fit_multiview(xList_train, y_train, xList_test, y_test, rho = 0.5)
```

### Prediction performance

Here, we outline the summary of predictive performance of cooperative
learning across different fusions.

```{r coop-results}
## View prediction performance results
multiview_summary <- data.frame(method = c("Early", "Late", "Intermediate"), mspe = c(fit_mv_early$mspe_mv, fit_mv_late$mspe_mv, fit_mv_intermediate$mspe_mv), nonzero_features = c(fit_mv_early$n_nonzero_mv, fit_mv_late$n_nonzero_mv, fit_mv_intermediate$n_nonzero_mv))
print(multiview_summary)
```

### Feature Selection Comparison

```{r coop-feature-selection}
## Extract coefficients for multiview, early fusion
nonzero_coef_mv_early <- fit_mv_early$nonzero_coefs

### Order features by coefficient value
nonzero_coef_mv_early_plot <- nonzero_coef_mv_early %>%
  arrange(coefficient) %>%
  mutate(feature = fct_reorder(feature, coefficient))

### Plot
p1 <- ggplot(nonzero_coef_mv_early_plot,
       aes(x = feature,
           y = coefficient,
           fill = view)) +
  geom_col(width = 0.7) +
  labs(
    title = "Non-Zero Regression Coefficients (Early Fusion)",
    x = "Feature",
    y = "Coefficient",
    fill = "View"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    plot.title = element_text(face = "bold")
  )

print(p1)

## Extract coefficients for multiview, intermediate fusion
nonzero_coef_mv_intermediate <- fit_mv_intermediate$nonzero_coefs

### Order features by coefficient value
nonzero_coef_mv_intermediate_plot <- nonzero_coef_mv_intermediate %>%
  arrange(coefficient) %>%
  mutate(feature = fct_reorder(feature, coefficient))

### Plot
p2 <- ggplot(nonzero_coef_mv_intermediate_plot,
       aes(x = feature,
           y = coefficient,
           fill = view)) +
  geom_col(width = 0.7) +
  labs(
    title = "Non-Zero Regression Coefficients (Intermediate Fusion)",
    x = "Feature",
    y = "Coefficient",
    fill = "View"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    plot.title = element_text(face = "bold")
  )

print(p2)

## Extract coefficients for multiview, late fusion
nonzero_coef_mv_late <- fit_mv_late$nonzero_coefs

### Order features by coefficient value
nonzero_coef_mv_late_plot <- nonzero_coef_mv_late %>%
  arrange(coefficient) %>%
  mutate(feature = fct_reorder(feature, coefficient))

### Plot
p3 <- ggplot(nonzero_coef_mv_late_plot,
       aes(x = feature,
           y = coefficient,
           fill = view)) +
  geom_col(width = 0.7) +
  labs(
    title = "Non-Zero Regression Coefficients (Late Fusion)",
    x = "Feature",
    y = "Coefficient",
    fill = "View"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    plot.title = element_text(face = "bold")
  )

print(p3)
```

## Bayesian Cooperative Learning using `BayesCOOP` R Package

While the `multiview` package implements cooperative learning by
combining prediction loss with an agreement penalty across multiple
views, it does **not inherently provide uncertainty quantification
(UQ)** for the fitted models or predictions. In many scientific
applications, especially those involving high stakes or interpretation
of effect sizes, simply reporting point estimates and predictive
accuracy is not sufficient; one often also needs credible intervals,
posterior distributions, or measures of confidence to assess the
reliability of findings.

`BayesCOOP` addresses this gap by placing a **Bayesian group
spike-and-slab double exponential (Laplace) prior** on the coefficients,
which encourages both sparsity and view-specific feature selection in a
principled probabilistic framework. It combines fast maximum a
posteriori (MAP) estimation with a **Bayesian Bootstrap** to approximate
posterior sampling and enable scalable uncertainty estimation for both
model parameters and predictions. This allows users to obtain credible
intervals, posterior summaries of view agreement parameters, and
measures of uncertainty that reflect uncertainty in the data and model
fit, rather than only point estimates.

Because of this Bayesian formulation, `BayesCOOP` not only selects
relevant features and captures cooperative effects across views but also
facilitates rigorous uncertainty quantification, which is important for
interpretation and for downstream decision-making in applications where
confidence in predictions matters. For further details, please refer to
the
[preprint](https://www.biorxiv.org/content/10.1101/2025.10.23.684056v1).

### Fit `BayesCOOP` (Bayesian Bootstrap Enabled)

Here we fit BayesCOOP using `bb = TRUE` to enable Bayesian
Bootstrap–based uncertainty estimation. This step may take a while
because it involves multiple bootstrap iterations, which can be adjusted
through the `bbiters` and `bbburn` parameters. The output includes the
posterior samples of the parameters and runtime.

```{r fit_bb_true}
set.seed(123)

## Fitting BayesCOOP
fit_BayesCOOP_bb_true <- BayesCOOP::BayesCOOP(data_train = data_train_processed,
  family = "gaussian",
  ss = c(0.05, 1),
  group = TRUE,
  bb = TRUE,
  alpha_dirich = 1,
  bbiters = 110,
  bbburn = 10,
  maxit = 100,
  warning = TRUE,
  verbose = FALSE,
  control = list()
)

```

### Fit `BayesCOOP` (Bayesian Bootstrap Disabled)

We also run BayesCOOP again with `bb = FALSE` to obtain posterior mode
estimates via a faster maximum a posteriori (MAP)-style variant (without bootstrap-based
uncertainty). All other settings remain unchanged.

```{r fit_bb_false}
set.seed(123)

## Fitting BayesCOOP with `bb` = FALSE
fit_BayesCOOP_bb_false <- BayesCOOP::BayesCOOP(
  data_train = data_train_processed,
  family = "gaussian",
  ss = c(0.05, 1),
  group = TRUE,
  bb = FALSE,
  alpha_dirich = 1,
  bbiters = 110,
  bbburn = 10,
  maxit = 100,
  warning = TRUE,
  verbose = FALSE,
  control = list(rho = 0.5)
)

## Prediction
pred_BayesCOOP_bb_false <- predict(object = fit_BayesCOOP_bb_false,
                                            newdata = data_test_processed, family = "gaussian",
                                            bb = FALSE, warning = TRUE, verbose = TRUE)

## Calculating mean squared prediction error
mspe_BayesCOOP_bb_false <- mean((pred_BayesCOOP_bb_false$y_pred - pred_BayesCOOP_bb_false$y_valid) ^ 2)
print(mspe_BayesCOOP_bb_false)
```

### Prediction Performance

```{r bayescoop-pp}
## Prediction when bb is TRUE
pred_BayesCOOP_bb_true <- predict(object = fit_BayesCOOP_bb_true,
                                            newdata = data_test_processed, family = "gaussian",
                                            bb = TRUE, warning = TRUE, verbose = TRUE)

## Calculating mean squared prediction error when bb is TRUE
mspe_BayesCOOP_bb_true <- mean((pred_BayesCOOP_bb_true$y_pred - pred_BayesCOOP_bb_true$y_valid) ^ 2)
print(mspe_BayesCOOP_bb_true)

## Prediction when bb is FALSE
pred_BayesCOOP_bb_false <- predict(object = fit_BayesCOOP_bb_false,
                                            newdata = data_test_processed, family = "gaussian",
                                            bb = FALSE, warning = TRUE, verbose = TRUE)

## Calculating mean squared prediction error when bb is FALSE
mspe_BayesCOOP_bb_false <- mean((pred_BayesCOOP_bb_false$y_pred - pred_BayesCOOP_bb_false$y_valid) ^ 2)
print(mspe_BayesCOOP_bb_false)

```

### Feature Selection / Sparsity Comparison

We now count the number of **selected (nonzero) features** for each
method to compare their **sparsity** and **selection aggressiveness**.

```{r pp_bcoop}
## Number of features selected for BayesCOOP with `bb` = TRUE
n_nz_BayesCOOP_bb_true  <- sum(fit_BayesCOOP_bb_true$beta_postmed!=0)
## Number of features selected for BayesCOOP with `bb` = FALSE
n_nz_BayesCOOP_bb_false <- sum(fit_BayesCOOP_bb_false$beta_MAP!=0)
## Number of features selected for multiview
sparsity_table <- rbind(
  data.frame(method = "BayesCOOP (bb=TRUE)",  n_selected_feat = n_nz_BayesCOOP_bb_true),
  data.frame(method = "BayesCOOP (bb=FALSE)", n_selected_feat = n_nz_BayesCOOP_bb_false)
)

## Print table comparing the number of selected features for diff methods
sparsity_table
```

Finally, we examine the features selected by `BayesCOOP`. For this
tutorial, we focus on the results obtained with `bb = TRUE`, since this
setting provides uncertainty quantification in addition to point
estimates, allowing us to assess both feature selection and associated
variability.

```{r feature_selection_bcoop}
# Posterior medians (already extracted)
beta_postmed <- fit_BayesCOOP_bb_true$beta_postmed
nonzero_pos  <- which(beta_postmed != 0)

coef_est <- beta_postmed[nonzero_pos]

# Posterior samples for selected coefficients
beta_postsamp <- fit_BayesCOOP_bb_true$beta_samples
postsamp_sel  <- beta_postsamp[, nonzero_pos]

# Compute credible intervals
q1 <- apply(postsamp_sel, 2, quantile, 0.025)
q3 <- apply(postsamp_sel, 2, quantile, 0.975)

# Create tidy dataframe
df_BayesCOOP <- data.frame(
  feature     = names(coef_est),
  coefficient = coef_est,
  q1          = q1,
  q3          = q3
)

df_BayesCOOP <- df_BayesCOOP %>%
  mutate(
    view = data_train_processed$feature_metadata$featureType[
      match(feature, data_train_processed$feature_metadata$featureID)
    ]
  ) %>%
  arrange(view, coefficient) %>%                         # optional: group ordering by view
  mutate(feature = fct_reorder(feature, coefficient))     # order bars by coefficient

p4 <- ggplot(df_BayesCOOP,
             aes(x = feature, y = coefficient, fill = view)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = q1, ymax = q3),
                width = 0.2,
                color = "black",
                linewidth = 0.6) +
  coord_flip() +
  labs(
    title = "BayesCOOP Selected Coefficients with 95% Credible Intervals",
    x = "Feature",
    y = "Coefficient",
    fill = "View"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold")
  )


print(p4)
```

## Factor Model Based Multiview Learning Approach using `jafar` R Package

Thus far, we integrated multiple views directly in the observed feature
space. Multiview factor models instead posit that each modality is
driven by a smaller set of latent factors, separating shared cross-view
signal from view-specific variation. This structured decomposition
provides a principled and interpretable framework for multimodal
integration.

The `jafar` package implements two supervised Bayesian factor modeling
approaches for multi-view data integration:

-   **Joint Factor Regression (**$\texttt{JFR}$): a baseline model that
    captures the combined variation across multiple data views using a
    single set of latent factors.
    $$\mathbf{x}_{mi} = \boldsymbol{\mu}_{m} + \boldsymbol{\Lambda}_{m}\boldsymbol{\eta}_{i} + \boldsymbol{\varepsilon}_{mi}\\y_{i} = \mu_{y} + \boldsymbol{\theta}^{\top}\boldsymbol{\eta}_{i} + e_{i}$$
-   **Joint Additive Factor Regression (**$\texttt{JAFAR}$): a more
    refined model that explicitly decomposes variation into shared and
    view-specific components. 
    $$\mathbf{x}_{mi} = \boldsymbol{\mu}_{m} + \boldsymbol{\Lambda}_{m}\boldsymbol{\eta}_{i} + \mathbf{\Gamma}_{m}\boldsymbol{\phi}_{mi}  + \boldsymbol{\varepsilon}_{mi}$$
    $$y_{i} = \mu_{y} + \Sigma_{m=1}^{M}\boldsymbol{\theta}^{\top}_{m}\boldsymbol{\phi}_{mi} + \boldsymbol{\theta}^{\top}\boldsymbol{\eta}_{i} + e_{i}$$
where $\mathbf{x}_{mi}\in \mathbb{R}^{p_{m}}, \mathbf{\Lambda}_{m}\in \mathbb{R}^{p_{m}\times K}, \boldsymbol{\eta}_{i}\in \mathbb{R}^{K}, \boldsymbol{\Gamma}_{m}\in \mathbb{R}^{p_{m}\times K_{m}}$ and $\boldsymbol{\phi}_{mi}\in \mathbb{R}^{K_{m}}$.

Isolating variation unique to each data modality, whether representing
genuine biological signal or measurement artifact, can be particularly
advantageous in high-dimensional settings. This separation enhances
interpretability and improves scalability.

$\texttt{JAFAR}$ enforces proper decomposition of additive components by
assigning zero prior mass to configurations that yield ill-posed
allocations of view-specific factors into the shared component.

Both models leverage extensions of the $\texttt{CUSP}$ prior
[(Legramanti et al.,
2020)](http://academic.oup.com/biomet/article/107/3/745/5847840),
learning adaptively the number of factors in a fully Bayesian way. The
package implements independent $\texttt{CUSP}$ priors
($\texttt{I-CUSP}$) for $\texttt{JFR}$ and dependent $\texttt{CUSP}$
priors ($\texttt{D-CUSP}$) for $\texttt{JAFAR}$.

In this tutorial, we focus on $\texttt{JAFAR}$ for ease of exposition.
Using $\texttt{JFR}$ is completely analogous, simply replacing the
function `gibbs_jafar()` with `gibbs_jfr()`\$.

### Preprocess the Data for `jafar` Implementation

```{r jafar-process}
## Calculate the number of views
views <- unique(data_train_processed$feature_metadata$featureType)
m <- length(views)

## pre-processing for jafar 
y <- data_train_processed$sample_metadata$Y
yTest <- data_test_processed$sample_metadata$Y

## Scaling the predictors for bmlasso
X_process <- lapply(xList_train, function(foo) as.matrix(scale(foo)))
X_process_test <- lapply(xList_test, function(foo) as.matrix(scale(foo)))

## centering the response
y_process <- y - mean(y)
y_process_test <- yTest - mean(yTest)
```

### Run `jafar`

```{r jafar, message=FALSE}
## Initialize Markov Chain Monte Carlo(MCMC) parameters for jafar
tMCMC = 10000
tBurnIn = 5000
tThin = 10
mcmc_seed=123
K0 = 25
K0_m = rep(20,m)

## Jafar Implementation
mcmc_supervised <- gibbs_jafar(X_process, y=y_process, K0=K0, K0_m=K0_m,
                               tMCMC=tMCMC, tBurnIn=tBurnIn, tThin=tThin,
                               hyperparams = list(seed = mcmc_seed))

```

### Prediction Performance

We can then compute response predictions and assess predictive
performance. Expected values and other distributional properties of
$y_i \mid \{X_{m i}\}_m$ can be easily obtained leveraging the latent
representation
$$ \mathbb{E}[\,y_i \mid \{X_{m i}\}_m \,] =  \mathbb{E}\big[\, \mathbb{E}[\,y_i \mid \eta_i, \{\phi_{m i}\}_m \,] \mid \{X_{m i}\}_m  \big]$$
This holds both for in-sample and out-of-sample observations.

```{r, message=FALSE}
## Prediction and mspe for jafar implemenation
y_pred <- predict_y(X_process_test, mcmc_supervised)
mspe_jafar <- mean((colMeans(y_pred$mean) - y_process_test)^2)
print(mspe_jafar)
```

### Some visualization
```{r jafar-vis}
## Plot the loading matrices
plot_loadings(mcmc_supervised, out_path = "~/Desktop/")
```

### Predictive Performance Comparison (MSPE) across different methods

The plot below summarizes the mean squared prediction error (MSPE)
across all methods evaluated in this lab for the StelzerDOS continuous
outcome dataset. Lower MSPE indicates better predictive accuracy on the
held-out test set.

```{r compare-mspe-barplot, message=FALSE, warning=FALSE}
# Compile all MSPEs into one tidy dataframe
mspe_all <- tibble(
  Method = c(
    "multiview (Early)",
    "multiview (Intermediate)",
    "multiview (Late)",
    "IntegratedLearner (Early)",
    "IntegratedLearner (Late)",
    "BayesCOOP (bb = TRUE)",
    "BayesCOOP (bb = FALSE)",
    "jafar"
  ),
  MSPE = c(
    multiview_summary %>% filter(method == "Early") %>% pull(mspe),
    multiview_summary %>% filter(method == "Intermediate") %>% pull(mspe),
    multiview_summary %>% filter(method == "Late") %>% pull(mspe),
    mspe_IL_early,
    mspe_IL_late,
    mspe_BayesCOOP_bb_true,
    mspe_BayesCOOP_bb_false,
    mspe_jafar
  )
)

# Create the barplot
ggplot(mspe_all, aes(x = fct_reorder(Method, MSPE), y = MSPE, fill = Method)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(MSPE, 3)), vjust = -0.5) +
  coord_flip() +
  labs(
    title = "Comparison of Predictive Performance (MSPE)",
    subtitle = "StelzerDOS continuous outcome",
    x = "Method",
    y = "Mean Squared Prediction Error"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold")
  )
```

# PRiSM Dataset (Binary Response)

While our main emphasis has been on continuous outcomes, the methodology
is not restricted to the Gaussian setting. We now briefly demonstrate
its application to a binary response.

To do so, we load the PRiSM dataset from the IntegratedLearner GitHub
repository, as described in the *IntegratedLearner* paper. This dataset
includes an independent **validation set**, enabling integrative
multi-omics modeling and rigorous out-of-sample evaluation.

In this dataset, we consider the same raw gut microbiome multi-omics dataset used in [Franzosa et al. (2019)](https://pubmed.ncbi.nlm.nih.gov/30531976/), derived from the Prospective Registry in IBD Study at MGH (PRISM). The study was designed to characterize gut microbial composition and metabolic potential in individuals with inflammatory bowel diseases (IBD).

Stool samples were collected from a cross-sectional discovery cohort of $155$ subjects, including $68$ individuals with Crohn’s disease (CD), $53$ with ulcerative colitis (UC), and $34$ non-IBD controls. The CD and UC groups are collectively categorized as IBD. A separate validation cohort of $65$ subjects was also considered for out-of-sample assessment.

Each stool sample underwent metagenomic sequencing, followed by profiling of microbial taxonomic composition and functional potential. After quality control, the dataset comprised a total of $9,171$ features across two molecular layers:

- **Metabolomics layer:** $8,831$ features  
- **Microbiome layer:** $340$ features  

The resulting dataset represents a high-dimensional multi-omics setting, with substantially more features than subjects. This structure motivates integrative modeling approaches capable of combining complementary biological signals across layers while controlling for overfitting in a large-\( p \), small-\( n \) regime.

Each dataset is structured as a list containing three coordinated
components that together define a multi-view data object:

-   **`feature_table`**: A matrix of high-dimensional molecular
    measurements derived from multiple omics platforms (e.g., microbiome
    profiles, metabolomics, proteomics, or other molecular assays). In
    this representation, **rows correspond to features** (such as taxa,
    metabolites, or proteins), and **columns correspond to samples**.
    This orientation is compatible with the modeling functions
    implemented in the IntegratedLearner workflow.

-   **`feature_metadata`**: A data frame providing feature-level
    annotations. These annotations describe the biological modality or
    data view (e.g., microbial and metabolite) This metadata enables
    modality-specific modeling, structured integration across views, and
    interpretation of feature importance.

-   **`sample_metadata`**: A data frame containing subject-level
    **binary outcome** information corresponding to the columns of
    `feature_table`.

```{r load_PRISM}
## Load data
data_train = get(load(url("https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/PRISM.RData"))); rm(pcl)
data_test = get(load(url("https://raw.githubusercontent.com/himelmallick/IntegratedLearner/master/data/NLIBD.RData"))); rm(pcl)

## How many layers and how many features per layer?
table(data_train$feature_metadata$featureType)

## Explore data dimension of feature_table: no. of features x no. of samples
print(dim(data_train$feature_table))

```

## Preprocessing the Dataset

As in the previous analysis, we rely on three internal utility functions
from the `BayesCOOP` package—`gen_datalist()`, `reconstruct_data()`, and
`filter_features()`—to carry out the preprocessing pipeline.

Specifically, `gen_datalist()` organizes the raw multi-view inputs into
a structured list format compatible with downstream modeling. The
`reconstruct_data()` function ensures consistent alignment of features
and samples across modalities, preserving correspondence between
predictors and outcomes. Finally, `filter_features()` performs
feature-level screening to remove low-quality or uninformative
variables, thereby stabilizing estimation and improving computational
efficiency.

Together, these functions standardize and streamline the data
preparation step prior to model fitting.

```{r PRISM_preprocess}
## Filtering features according to abundance and prevalence threshold
xList_train <- BayesCOOP:::gen_datalist(data_train)$xList
y_train <- BayesCOOP:::gen_datalist(data_train)$y
xList_test <- BayesCOOP:::gen_datalist(data_test)$xList
y_test <- BayesCOOP:::gen_datalist(data_test)$y
abd_thresh <- 0; prev_thresh <- 0.1
xList_train <- lapply(xList_train, function(foo) BayesCOOP:::filter_features(foo, abd_thresh, prev_thresh))
xList_test <- lapply(xList_test, function(foo) BayesCOOP:::filter_features(foo, abd_thresh, prev_thresh))

## Considering the common features between train and test set
keep_features <- vector("list", length = length(xList_train))
for(i in 1:length(keep_features)) {
  keep_features[[i]] <- intersect(names(xList_train[[i]]), names(xList_test[[i]]))
  xList_train[[i]] <- as.matrix(xList_train[[i]][, keep_features[[i]], drop = FALSE])
  xList_test[[i]] <- as.matrix(xList_test[[i]][, keep_features[[i]], drop = FALSE])
}

## Reconstructing the train and testing data object
data_train_processed <- BayesCOOP:::reconstruct_data(xList_train, y_train)
data_test_processed <- BayesCOOP:::reconstruct_data(xList_test, y_test)

# Change column name to make it uniform across all packages henceforth
colnames(data_train_processed$sample_metadata)[1] <- "subjectID"
colnames(data_train_processed$feature_metadata) <- c("featureID", "featureType")
colnames(data_test_processed$sample_metadata)[1] <- "subjectID"
colnames(data_test_processed$feature_metadata) <- c("featureID", "featureType")

## Check new dimensions after filtering of features
print(dim(data_train_processed$feature_table))
print(dim(data_test_processed$feature_table))
```

While the `IntegratedLearner` and `multiview` packages currently support
binary outcomes, the `jafar` and `BayesCOOP` packages do not yet provide
this functionality. We are actively working toward extending these
frameworks to accommodate binary responses in future releases.

## Integrated Learner using `IntegratedLearner` R Package

```{r PRISM_IL, message=FALSE, warning=FALSE}
## Fit IntegratedLearner
fit_IntegratedLearner <- IntegratedLearner(
      feature_table = data_train_processed$feature_table,
      sample_metadata = data_train_processed$sample_metadata,
      feature_metadata = data_train_processed$feature_metadata,
      feature_table_valid = data_test_processed$feature_table,
      sample_metadata_valid = data_test_processed$sample_metadata,
      folds = 5,
      seed = 123,
      base_learner = "SL.randomForest",
      base_screener = "All",
      meta_learner = "SL.nnls.auc",
      run_concat = TRUE,
      run_stacked = TRUE,
      verbose = FALSE,
      print_learner = TRUE,
      refit.stack = FALSE,
      family = binomial()
    )

```

### Prediction Performance

```{r, PRISM_IL_pp}
## Prediction values of Y for both early and late fusion schemes
y_pred_early <- fit_IntegratedLearner$yhat.test$concatenated
y_pred_late <- fit_IntegratedLearner$yhat.test$stacked
```

`IntegratedLearner` offers easily accessible and interpretable summary
outputs including 1) computation time, 2) per-layer AUC/ $R^2$ scores on
training and/or test data (if `feature_table_valid` and
`sample_metadata_valid` are provided), 3) AUC/ $R^2$ metrics for stacked
and concatenated model if `run_stacked=TRUE` and `run_concat=TRUE` and
4) estimated per-layer weights from meta learner in stacked model.

We can visualize the classification performance by constructing
layer-wise ROC curves for both train and test set using `plot.learner()`
function that takes `IntegratedLearner` object as input:

```{r, warning=FALSE, fig.height=10, fig.width=7}
plot.obj <- IntegratedLearner:::plot.learner(fit_IntegratedLearner)
```

## Cooperative Learning using `multiview` R Package

```{r PRISM_coop}
## Matrix datatype conversion
xList_train <- lapply(xList_train, as.matrix)
xList_test <- lapply(xList_test, as.matrix)

## Different rho values for choice of fusion schemes
rho = 0.5 # Intermediate Fusion

## Run cross-validation to choose the best lambda (regularization strength)
cvfit <- multiview::cv.multiview(x_list = xList_train, y = y_train, rho = rho, alpha = 1, nfolds = 5)
## Plot the cross-validation curve
plot(cvfit)
## Extract the CV-selected optimal lambda (often minimizing CV error)
best_lambda <- cvfit$lambda_min

## Fit the multiview model on the training data 
fit_mv <- multiview::multiview(
  xList_train, y_train,
  lambda = best_lambda,
  rho = rho,
  alpha = 1,
  family = binomial()
)

## Predict responses on the test set.
y_pred_mv <- predict(
  fit_mv,
  newx = xList_test,
  s = best_lambda,
  rho = rho,
  alpha = 1
)

## Extract fitted coefficients at the selected lambda.
coefs_mv <- coef(
  fit_mv,
  s = best_lambda,
  rho = rho,
  alpha = 1
)

## Count how many coefficients are nonzero, then subtract 1 to exclude intercept.
n_nonzero_mv <- sum(rowSums(coefs_mv != 0) > 0) - 1  # Exclude the intercept
print(n_nonzero_mv)
```

## Looking Ahead

Looking ahead, **IntegratedLearner (IL)** is being expanded to accommodate survival and multiclass outcomes, broadening its applicability to time-to-event analyses and complex phenotypic classifications that are common in modern biomedical studies. These extensions will preserve IL’s uncertainty-aware design while enabling principled inference beyond binary and continuous endpoints.

Together, **BayesCOOP**, **JAFAR**, and **IntegratedLearner** can be viewed as *Bayesian sisters* - each addressing a distinct fusion paradigm (intermediate, factor-based, and ensemble-based integration, respectively) within a coherent uncertainty quantification framework. In contrast, **multiview (cooperative learning)** serves as the frequentist cousin, conceptually connecting early, intermediate, and late fusion strategies across these paradigms.

This evolving ecosystem provides a unified statistical language for multimodal data integration, balancing interpretability, scalability, and rigorous uncertainty quantification across diverse outcome types.