---
title: "Module 2: Horizontal Integration (Meta-Analysis)"
author: "Sreya Sarkar, University of Iowa"
date: "ISMB 2026, Washington DC" 
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    theme: readable
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    output_dir: '.'
  header-includes:
  - \DeclareUnicodeCharacter{2212}{-}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

> **ISMB 2026 Tutorial VT4, Multimodal Integration and Multimodal Causal Inference using R/Bioconductor.** This is Module 2 of three hands-on modules delivered in a single day. Having covered vertical (multi-omics) integration in Module 1, we now turn to horizontal integration, combining evidence across studies that share a common feature space. Module 3 will close the day with multiview mediation using Zentangler. If you have not already done so, run the setup script once (`Rscript setup/install.R`) to install and load every package used across all three modules.

## Setup

Install the following R packages to get started.

```{r R-libs, message = FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Global Packages for data availability and/or visualization/demonstration
library(curatedMetagenomicData) ## source: Bioconductor
library(cowplot) ## source: CRAN
library(ggplot2) ## source: CRAN
library(ggpubr) ## source: CRAN
library(dplyr) ## source: CRAN
library(tidyverse) ## source: CRAN
library(parallel) ## Base R package

## For MMUPHin
library(vegan) ## source: CRAN
library(MMUPHin) ## source: Bioconductor

## For Network visualization
library(SpiecEasi) ## source: Github, devtools::install_github("zdk123/SpiecEasi")
library(SPRING) ## source: Github, devtools::install_github("GraceYoon/SPRING")
library(igraph) ## source: CRAN
library(NetCoMi)
## source: Github
# devtools::install_github("stefpeschel/NetCoMi", 
#                          repos = c("https://cloud.r-project.org/",
#                                    BiocManager::repositories()))

## For muffinette
library(muffinette) # source: Github, devtools::install_github("himelmallick/muffinette")
```


Meta-analysis, a form of horizontal integration, is a statistical analysis
that combines the results of multiple scientific studies by appropriately
weighting the individual study-specific effect sizes. A common meta-analytic
combination strategy is to take a weighted average of the study-specific
summary measures. In fixed-effects meta-analysis, the weights are based on
the assumption that there is a single true parameter underlying all the studies,
while in random-effects meta-analysis, the weights are based on a model where
the true parameter varies across studies according to a probability
distribution.

Meta-analysis is particularly useful when individual studies are small,
and a joint analysis of these studies can enhance the precision of
effect size estimation and improve statistical power. Especially in
microbiome studies, individual effect sizes from each study tend to be
small. Only through meta-analysis can these small effects reach
statistical significance. Therefore, by jointly analyzing these studies,
we amplify the collective voice of the data, offering a more
comprehensive understanding of the effects under investigation.

In this discussion, we demonstrate the implementation of horizontal integration
by utilizing publicly available cancer microbiome datasets from the
R/Bioconductor package `curatedMetagenomicData`. Our demonstration comprises
three components as follows:

- [`MMUPHin`](https://www.bioconductor.org/packages/release/bioc/html/MMUPHin.html)(Meta-Analysis Methods with a Uniform Pipeline for Heterogeneity in microbiome studies) - [A Unimodal Framework]
- [`muffinette`](https://github.com/himelmallick/muffinette)(Meta-Analysis of Differential Microbiome Networks) - [A Unimodal Framework]


## Meta-analysis of cancer microbiome studies using relative abundance data (MMUPHin)

In this section, we provide detailed examples of how to perform
batch effect correction and meta-analytic differential abundance testing
on publicly available cancer microbiome data from multiple studies using the
R package `MMUPHin`. `MMUPHin` is a R/Bioconductor package designed for
meta-analysis of microbiome taxonomic and functional profiles. It is agnostic
to the data type and leverages another R/Bioconductor package, `MaAsLin2`, as a
backend to conduct the meta-analysis. `MaAsLin2` is particularly
designed to be applicable to various microbial community data types
(taxonomy or functional profiles) and environments (human or otherwise).
It is modular, including implementations of alternative
normalization/transformation schemes and statistical models (e.g., for
amplicon vs. shotgun metagenomic profiles). Leveraging this flexible
framework under the hood, `MMUPHin` performs normalization, batch effect
correction, and meta-analysis using the default random effects
meta-regression.


### Data Pre-processing 

We use both the relative abundance data and the pathway abundance data from 5
public datasets. First, let us look into the relative abundance data from the
`curatedMetagenomicData` package. In this analysis, we will use two
input files. The first input file is the relative abundance data file, which
consists of **a feature-by-sample matrix**. The second input is
**the metadata** associated with the relative abundance file, which includes the
study names and combines overall response rate (ORR), progression-free survival
at 12 months (PFS12), as well as the response to anti-PD-1 (anti_PD_1) as the
response variable.


```{r load-pkg-data, message=FALSE, warning=FALSE}

######################
# Load melanoma data #
######################

se_relative <- sampleMetadata |>
  filter(study_name %in% c("FrankelAE_2017", "GopalakrishnanV_2018", "LeeKA_2022", 
                         "MatsonV_2018", "PetersBA_2019", "WindTT_2020")) |>
  returnSamples("relative_abundance", rownames = "short")

```


```{r prepare-input-data}

##########################
# Create sample metadata #
##########################

data_meta <- dplyr::select(as.data.frame(colData(se_relative)),
                    c("study_name", "ORR", "PFS12", "anti_PD_1"))

# Define response variable
data_meta$resvar <- NA
data_meta$anti_PD_1[data_meta$anti_PD_1 == "responder"] <- "yes"
data_meta$anti_PD_1[data_meta$anti_PD_1 == "non_responder"] <- "no"
data_meta$resvar[!is.na(data_meta$anti_PD_1)] <- data_meta$anti_PD_1[!is.na(data_meta$anti_PD_1)]
data_meta$resvar[!is.na(data_meta$PFS12)] <- data_meta$PFS12[!is.na(data_meta$PFS12)]
data_meta$resvar[!is.na(data_meta$ORR)] <- data_meta$ORR[!is.na(data_meta$ORR)]


cat(sprintf("Before filtering samples: Total %d samples from %d studies.\n\n",
            nrow(data_meta), length(unique(data_meta$study_name))))

# Filter individuals without response variable
data_meta <- data_meta[!is.na(data_meta$resvar), c("study_name", "resvar")]
cat(sprintf("After filtering samples: Total %d samples from %d studies.\n\n",
            nrow(data_meta), length(unique(data_meta$study_name))))


# Convert the "study_name" to factor variable
data_meta$study_name <- as.factor(data_meta$study_name)


###########################
# Create Species Features #
###########################
# Transpose the abundance matrix and change the value of abundance
# data to proportion unit
data_abd <- assay(se_relative)
data_abd <- data_abd / 100

# Match the individuals in the data_abd
data_abd <- data_abd[, colnames(data_abd) %in% rownames(data_meta)]

# Change the rownames names for variables of interest
rownames(data_abd) <- sub('.*s__', '', rownames(data_abd))
```


To reduce noise and remove uninformative features, we filtered the data
using predefined **abundance** and **prevalence** thresholds. Abundance
filtering removes consistently negligible features, while prevalence
filtering excludes features present in only a small fraction of samples.

In our analysis, we set the abundance threshold to 0, meaning no
features were removed based on magnitude, and applied a prevalence
threshold of 0.1, retaining features observed in at least 10% of
samples. This reduces instability from extremely rare features while
improving interpretability and computational efficiency. For this purpose,
we use the `filter_abdfeatures()` function from the R package `muffinette`.

Next, we apply an additional step of filtering to remove features exhibiting
little variability across samples. To achieve this,
we utilize the `filter_varfeatures()` function from the R package `muffinette`.
This function first eliminates the features with near-zero variance, and sorts
the remaining features in decreasing order of variance. For computational
efficiency, this function further allows the user to retain only
the top $K$ most variable features.


```{r study-wise-filtering-features}
###############################
# Study-wise feature filtering #
###############################

# Make sure abundance matrix columns are exactly the filtered samples
# and in the same order as data_meta
data_abd <- data_abd[, rownames(data_meta), drop = FALSE]

# Optional sanity check
stopifnot(all(colnames(data_abd) == rownames(data_meta)))

# Split sample IDs by study
samples_by_study <- split(rownames(data_meta), data_meta$study_name)

# Apply filtering study-wise
data_abd_filtered_list <- lapply(names(samples_by_study), function(study) {
  
  samp_ids <- samples_by_study[[study]]
  
  # Feature-by-sample matrix for one study
  abd_study <- data_abd[, samp_ids, drop = FALSE]
  
  cat(sprintf(
    "\nStudy: %s | Before filtering: %d samples on %d features.\n",
    study, ncol(abd_study), nrow(abd_study)
  ))
  
  # filter_abdfeatures expects sample-by-feature matrix
  abd_study_filtered <- filter_abdfeatures(
    x = t(abd_study),
    abd_threshold = 0,
    prev_threshold = 0.1)
  
  # filter variable features within the same study
  abd_study_filtered <- filter_varfeatures(
    x = abd_study_filtered,
    topV = NULL)
  
  # Transpose back to feature-by-sample matrix
  abd_study_filtered <- as.matrix(t(abd_study_filtered))
  
  cat(sprintf(
    "Study: %s | After filtering: %d samples on %d features.\n",
    study, ncol(abd_study_filtered), nrow(abd_study_filtered)
  ))
  
  return(abd_study_filtered)
})

names(data_abd_filtered_list) <- names(samples_by_study)
str(data_abd_filtered_list)

## Pooled feature-by-sample matrix
common_features <- Reduce(intersect, lapply(data_abd_filtered_list, rownames))

data_abd_filtered_common_list <- lapply(data_abd_filtered_list, function(x) {
  x[common_features, , drop = FALSE]
})
data_abd_filtered_common <- do.call(cbind, data_abd_filtered_common_list)

dim(data_abd_filtered_common) 

data_abd_filtered_common <- data_abd_filtered_common[, rownames(data_meta), drop = FALSE]

stopifnot(all(colnames(data_abd_filtered_common) == rownames(data_meta)))
```



At the end of these steps, we have prepared the two required input files
consisting of **285 samples of relative abundances on 121 features collected
from 5 different studies**.



### Performing batch (study) effect adjustment with adjust_batch for relative abundance {#sec-batch-effect-adjustment}

Here, we will perform the batch effect correction analysis using the
`adjust_batch` function from `MMUPHin`, adjusting for the response
variable `resvar`. A brief description of the function
arguments is as follows:

- `feature_abd`: feature-by-sample matrix of abundances (proportions/counts)
- `batch`: name of the batch variable (should be a factor variable included in
the metadata)
- `covariates`: names of covariates to adjust for in the batch-correction model
- `data`: data frame of metadata, columns must include exposure and batch and
covariates to adjust for (if specified)


```{r batch-effect-corret}
# Use adjust_batch to correct for differences in the five studies, while
# controlling for the effect of cases versus control (variable resvar in data_meta).
fit_adjust_batch <- MMUPHin::adjust_batch(feature_abd = data_abd_filtered_common,
                                 batch = "study_name",
                                 covariates = "resvar",
                                 data = data_meta,
                                 control = list(verbose = FALSE))

data_abd_adj <- fit_adjust_batch$feature_abd_adj

```

By the end of this step, we obtain a feature-by-sample
(`r dim(data_abd_adj)[1]` $\times$ `r dim(data_abd_adj)[2]`) batch-adjusted
relative abundance matrix.



### Evaluation for batch effect adjustment {#sec-batch-adjustment-evaluation}

After obtaining the adjusted abundance matrix, we can evaluate the
improvement of the batch effect adjustment. Here, the total variation
from study difference will be assessed through the PERMANOVA$^14$ test, using
the R package `vegan`.

```{r permanovaTest}

D_before <- vegan::vegdist(t(data_abd_filtered_common))
D_after <- vegan::vegdist(t(data_abd_adj))

set.seed(1)
fit_adonis_before <- vegan::adonis2(D_before ~ study_name, data = data_meta)
fit_adonis_after <- vegan::adonis2(D_after ~ study_name, data = data_meta)

print(fit_adonis_before)
print(fit_adonis_after)

```

Based on the results, we can see that the study differences can explain
a total of `r round(fit_adonis_before$R2[1]*100, 1)`% ($R^2$) of the variability
in microbial abundance profiles before study effect adjustment, whereas after
adjustment this was reduced to `r round(fit_adonis_after$R2[1]*100, 1)`% ($R^2$). 

Let us now visualize the results of the PERMANOVA test.

```{r visualization-Permanova}

##############
# Ordination #
##############
# Before
R2_before <- round(fit_adonis_before$R2[1]*100, 1)
pcoa_before <- cmdscale(D_before, eig = TRUE)
ord_before <- as.data.frame(pcoa_before$points)
percent_var_before <- round(pcoa_before$eig / sum(pcoa_before$eig) * 100, 1)[1:2]
before_labels <- c(paste('Axis 1 (', percent_var_before[1], '%)', sep = ''), 
                 paste('Axis 2 (', percent_var_before[2], '%)', sep = ''))

before_phrase <- paste('Before (PERMANOVA R2 = ', R2_before, '%)', sep = '')
colnames(ord_before) <- c('PC1', 'PC2')
ord_before$Study <- data_meta$study_name

# After
R2_after <- round(fit_adonis_after$R2[1]*100, 1)
pcoa_after <- cmdscale(D_after, eig = TRUE)
ord_after <- as.data.frame(pcoa_after$points)
percent_var_after <- round(pcoa_after$eig / sum(pcoa_after$eig) * 100, 1)[1:2]
after_labels <- c(paste('Axis 1 (', percent_var_after[1], '%)', sep = ''), 
                paste('Axis 2 (', percent_var_after[2], '%)', sep = ''))

after_phrase <- paste('After (PERMANOVA R2 = ', R2_after, '%)', sep = '')
colnames(ord_after) <- c('PC1', 'PC2')
ord_after$Study <- data_meta$study_name

# Ordination Plot
p_before <- ggplot(ord_before, aes(x = PC1, y = PC2, color = Study)) + 
  geom_point(size = 4) + 
  theme_bw() + 
  xlab(before_labels[1]) + 
  ylab(before_labels[2]) + 
  ggtitle(before_phrase) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(legend.position ="none")

p_after <- ggplot(ord_after, aes(x = PC1, y = PC2, color = Study)) + 
  geom_point(size = 4) + 
  theme_bw() + 
  xlab(after_labels[1]) + 
  ylab(after_labels[2]) + 
  ggtitle(after_phrase) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(legend.position ="none")

p <- plot_grid(p_before, p_after, ncol = 2)
p

```



### Meta-analytical differential abundance testing with maaslin_meta {#sec-differential-abundance-testing}

Here, we illustrate the details of the meta-analytical differential
abundance testing through the `maaslin_meta` function provided by the `MMUPHin`
packages with the following specifications:

- `feature_abd`: feature-by-sample matrix of abundances (proportions/counts)
- `batch`: name of the batch variable (should be a factor variable included in
the metadata)
- `exposure`: name of the exposure variable for differential abundance testing
- `covariates`: names of covariates to adjust for in `MaAslin2` differential
abundance testing models.
- `covariates_random`: names of random-effects grouping covariates to adjust
for in `MaAslin2` differential abundance testing models.
- `data`: data frame of metadata, columns must include exposure and batch and
covariates to adjust for (if specified)
- `control`: a named list of additional parameters including normalization
technique, transformation to use and analysis method. See `MaAsLin2` for details.
Moreover, several choices of estimation method are available for fitting either a
fixed-effects or a random-effects model for meta-analysis. Refer to `rma.uni()` in
the R package `metafor`.

Here, we apply the Tweedie model using the "CPLM" analysis
method in `MaAsLin2` and pass it as part of control. 

```{r Implement-mmuphin-DAtest, warning=FALSE}
fit_maaslin_meta <- maaslin_meta(feature_abd = data_abd_adj,
                       batch = "study_name",
                       exposure = "resvar",
                       data = data_meta,
                       control = list(analysis_method = 'CPLM', transform = 'LOG'))

meta_fits_mmuphin_abundance <- fit_maaslin_meta$meta_fits_abundance
meta_fits_mmuphin_prevalence <- fit_maaslin_meta$meta_fits_prevalence

signif_species_meta_abundance <- meta_fits_mmuphin_abundance[meta_fits_mmuphin_abundance$qval.fdr < 0.05, 1]
signif_species_meta_prevalence <- meta_fits_mmuphin_prevalence[meta_fits_mmuphin_prevalence$qval.fdr < 0.05, 1]
signif_species_meta_abundance
signif_species_meta_prevalence
```

This step yields a data frame of per-feature meta-analytical differential abundance
results, including columns for effect sizes, p-values and q-values, heterogeneity statistics such as $\tau^2$ and $I^2$, as well as weights for individual batches.
More details of these statistics are included in `rma.uni()` in `metafor` package.


Based on the results from `MMUPHin`, we observe that there are NO significant
species.


## Meta-network-analysis (muffinette)

Microbial communities operate through complex ecological interactions that are
better understood via network-based analyses. Many existing work in the
literature (eg. SparCC, SpiecEasi etc.) construct microbial association networks
within individual studies in order to capture interspecies relationships.
`muffinette` is a novel statistical framework designed to meta-analyze such
networks across studies to identify reproducible interaction patterns.

After the pre-processing steps similar to those used in `MMUPHin`, including
prevalence and variance-based filtering, and followed by adjustment of batch
(study) effects, `muffinette` introduces the following additional steps of
network-level connectivity modeling procedures, executed per study, prior to
conducting meta-analysis across studies.

(1) Estimation of within-study correlation networks
(2) Calculation of feature-level connectivity scores
(3) Calculation of feature-by-sample pseudovalue matrix

Here, we demonstrate these steps for one study.

```{r prepare-per-study-abd}
batch_levels <- levels(as.factor(data_meta$study_name))
exp_levels <- levels(as.factor(data_meta$resvar)) 

## vector of sample sizes for different studies
ni <- as.numeric(table(factor(data_meta$study_name, levels = batch_levels)))
ni_ends <- cumsum(ni)
ni_starts <- c(1, ni_ends[-length(ni)] + 1)

## Consider the first study for demonstration
## Extract the corresponding batch-corrected abundance matrix
study1 <- batch_levels[1]
data_abd_adj_all <- t(data_abd_adj)
study1_abd <- data_abd_adj_all[ni_starts[1]:ni_ends[1], ]

## extracting indices of two groups of response (No and Yes)
groupA <- which(data_meta$resvar[data_meta$study_name == study1] == exp_levels[1])
groupB <- which(data_meta$resvar[data_meta$study_name == study1] == exp_levels[2])


## extracting the group-specific abundances
xA <- study1_abd[groupA, , drop = FALSE]
xB <- study1_abd[groupB, , drop = FALSE]


## Remove the study from analysis if sample size is limited
if(nrow(xA) < 5 | nrow(xB) < 5) {
    message(sprintf("Skipping study %s: too few samples per
    exposure (nA = %d, nB = %d)", studies[i], nrow(xA), nrow(xB)))
    estimatedNet <- NULL
    pseudoVal <- NULL
}
```


```{r network-steps}
fixseed <- 1
ncores <- 4

## Setting seed for reproducibility across parallel network estimations
cl <- parallel::makeCluster(ncores)
# on.exit(parallel::stopCluster(cl), add = TRUE)
if(!is.null(fixseed)) {
# ensure master uses reproducible parallel RNG
RNGkind("L'Ecuyer-CMRG")
set.seed(fixseed)

# ensure workers use same RNG engine
parallel::clusterEvalQ(cl, {
    RNGkind("L'Ecuyer-CMRG")
    NULL
})

# generate independent streams for workers
parallel::clusterSetRNGStream(cl, iseed = fixseed)
}

parallel::clusterExport(cl, varlist = c("networkEst"),
                envir = environment())


## Estimation of feature-by-feature association matrix
estimatedNet_groupA <- networkEst(x = xA,
                                  estimethod = "SparCC",
                                  iter = 20, inner_iter = 10, th = 0.1)
estimatedNet_groupB <- networkEst(x = xB,
                                  estimethod = "SparCC",
                                  iter = 20, inner_iter = 10, th = 0.1)
estimatedNet <- list(estimatedNet_groupA = estimatedNet_groupA,
                     estimatedNet_groupB = estimatedNet_groupB)


## global network-connectivity measurement
thetahat_groupA <- measureNetwork(estimatedNet_groupA)
thetahat_groupB <- measureNetwork(estimatedNet_groupB)


## Estimation of feature-by-feature association matrix
## leaving one-subject-out
## using parLapply for computational efficiency
clusterExport(cl, varlist = c("xA"),
                        envir = environment())

estimatedNet_loo_groupA <- parLapply(cl, X = seq_len(nrow(xA)),
            fun = function(j) {
                tryCatch(networkEst(x = xA[-j, , drop = FALSE],
                        estimethod = "SparCC",
                        iter = 20, inner_iter = 10, th = 0.1),
                            error = function(e) e)})
clusterExport(cl, varlist = c("xB"),
                        envir = environment())

estimatedNet_loo_groupB <- parLapply(cl, X = seq_len(nrow(xB)),
            fun = function(j) {
                tryCatch(networkEst(x = xB[-j, , drop = FALSE],
                        estimethod = "SparCC",
                        iter = 20, inner_iter = 10, th = 0.1),
                            error = function(e) e)})        

## global network-connectivity measurement leaving one-subject-out
thetahat_loo_groupA <- sapply(estimatedNet_loo_groupA, measureNetwork)
thetahat_loo_groupB <- sapply(estimatedNet_loo_groupB, measureNetwork)

parallel::stopCluster(cl)
rm(cl)
```


Let us visualize these two networks from the two groups
using the `NetCoMi` package.

```{r visualize-networks}
net <- netConstruct(data = as.matrix(estimatedNet_groupA),
                    data2 = as.matrix(estimatedNet_groupB),
                    dataType  = "correlation", 
                    sparsMethod = "threshold",
                    dissFunc = "unsigned",
                    thresh = 0.19,
                    weighted  = TRUE,
                    verbose = 0)

A1 <- net$assoMat1 
A2 <- net$assoMat2 

g1 <- graph_from_adjacency_matrix(A1, mode="undirected", weighted=TRUE, diag=FALSE)
g2 <- graph_from_adjacency_matrix(A2, mode="undirected", weighted=TRUE, diag=FALSE)

# Layout graph with positive weights (abs), but keep original g1/g2 for plotting
g1_layout <- g1
E(g1_layout)$weight <- abs(E(g1_layout)$weight)

L <- layout_with_fr(g1_layout) 

# edge aesthetics using ORIGINAL sign
E(g1)$color <- ifelse(E(g1)$weight > 0, "darkgreen", "red")
E(g2)$color <- ifelse(E(g2)$weight > 0, "darkgreen", "red")
E(g1)$width <- 1 + 4*abs(E(g1)$weight)
E(g2)$width <- 1 + 4*abs(E(g2)$weight)


par(mfrow=c(1,2), mar=c(1,1,2,1))
plot(g1, layout=L, vertex.size=6, vertex.label=NA, main="Group 1 (No)")
plot(g2, layout=L, vertex.size=6, vertex.label=NA, main="Group 2 (Yes)")

# move to outer margin space
par(xpd=NA)

legend(x = -2.72, y = -1.15,
       legend = c("Positive association", "Negative association"),
       col = c("darkgreen", "red"),
       lwd = 3,
       horiz = TRUE,
       inset = -0.001,
       bty = "n",
       cex = 1)
par(mfrow=c(1,1), xpd=FALSE)
```




We now calculate the jackknife pseudovalues for the $i$-th subject and $k$-th
feature as

$$\tilde{\theta}_{ik} = n\hat{\theta}_k - (n-1)\hat{\theta}_{k(-i)}$$.

We use the `pseudoValue` function from the `muffinette` package to do this task.
The function takes the following arguments:

* `thetahatinput` network measure of all subjects,
* `thetahatdropinput` network measure leaving-one-subject-out,
* `sizegroup` number of observations,

and returns a sample-by-feature matrix of jackknife pseudovalues.


```{r pseudovalues}
## Calculation of jackknife pseudovalues
pseudoVal_groupA <- pseudoValue(thetahat_groupA, thetahat_loo_groupA, nrow(xA))
pseudoVal_groupB <- pseudoValue(thetahat_groupB, thetahat_loo_groupB, nrow(xB))
pseudoVal <- matrix(NA, nrow(study1_abd), ncol(study1_abd))
pseudoVal[groupA, ] <- pseudoVal_groupA
pseudoVal[groupB, ] <- pseudoVal_groupB
pseudoVal <- scale(pseudoVal, center = TRUE, scale = TRUE)

## Maps the column names (feature names)
colnames(pseudoVal) <- colnames(study1_abd)

cat(sprintf("A sample-by-feature (%d x %d) matrix of
jackknife pseudovalues for study %s:\n\n", nrow(pseudoVal),
ncol(pseudoVal), study1))
print(pseudoVal[1:4, 1:3])

```

We repeat the aforesaid steps for every study and construct a sample-by-feature
(`r dim(data_abd_adj)[2]`$\times$ `r dim(data_abd_adj)[1]`) pseudovalue matrix.
Finally, we perform meta-network-analysis of these pseudovalues across studies
to discover consistent network-based signatures. The `muffinette` function in
the R package `muffinette` implements this entire workflow as demonstrated below.

```{r Implement-muffinette}
###########################################
# Study-wise abundance list without filter #
###########################################

# Make sure abundance matrix columns are exactly the filtered samples
# and in the same order as data_meta
data_abd <- data_abd[, rownames(data_meta), drop = FALSE]

# Sanity check
stopifnot(all(colnames(data_abd) == rownames(data_meta)))

# Split sample IDs by study
samples_by_study <- split(rownames(data_meta), data_meta$study_name)

# Create study-wise feature-by-sample abundance matrices
data_abd_list <- lapply(names(samples_by_study), function(study) {
  
  samp_ids <- samples_by_study[[study]]
  
  abd_study <- data_abd[, samp_ids, drop = FALSE]
  
  cat(sprintf(
    "\nStudy: %s | %d samples on %d features.\n",
    study, ncol(abd_study), nrow(abd_study)
  ))
  
  return(abd_study)
})

names(data_abd_list) <- names(samples_by_study)

str(data_abd_list)

fit_muffinette <- muffinette(metaAbd = data_abd,
                                    feature_abd_list = data_abd_list,
                                    batchvar = data_meta$study_name,
                                    exposurevar = data_meta$resvar,
                                    metaData = data_meta,
                                    prevfilter = TRUE, filter_perstudy = TRUE,
                                    topfeatures = 100,
                                    comp = TRUE, net.est.method = "SparCC",
                                    covariates = NULL, covariates_random = NULL,
                                    ncores = 4, verbose = TRUE, fixseed = NULL, control = NULL)

meta_fits_muffinette <- fit_muffinette$metafits
View(meta_fits_muffinette)
signif_species_muffinette <- meta_fits_muffinette[meta_fits_muffinette$qval < 0.05, 1]
```


Let us now evaluate the results obtained by `muffinette`.

```{r comparison}
fig_muffinette <- meta_fits_muffinette %>%
  filter(qval < 0.05) %>%
  mutate(feature = sub("^species:", "", feature)) %>%
  ggplot(aes(y = coef, x = reorder(feature, coef))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 5)) +
  labs(y = "Estimated Coefficients", x = "Differentially Abundant Species", title = "muffinette")


# fig <- ggarrange(fig_mmuphin, fig_muffinette, nrow = 2, heights = c(0.75, 1))
# fig <- annotate_figure(fig,
#                        left = text_grob("Differentially Abundant Species",
#                                         size = 8, face = "bold", rot = 90),
#                        bottom = text_grob("Estimated Coefficients",
#                                           size = 8, hjust = 0.5, face = "bold"))
fig_muffinette
```



* `muffinette` identifies 8 significant species in total, among which
\emph{Phocaeicola dorei} has the strongest positive effect, whereas
\emph{Collinsella aerofaciens} has the strongest negative effect. These features
could only be detected by meta-analysis of the network-relationships within a
microbiome community unlike `MMUPHin` which directly meta-analyzes relative
abundances.






## Session information

```{r}
sessionInfo()
```

